Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-05-07 analysis/paper/paper.qmd
editing with peer reviews

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades. This study presents a bibliometric assessment of archaeology’s status relative to other fields, alongside a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. PC1 is explained by title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. PC2 is explained by number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of PC2 are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as generalist communities of practice.

>>>>

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. Variation in PC2 is influenced by the number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.
